Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-09-30 chapters/chapter01.tex
beast subsection

\subsection{Bayesian evolutionary inference in BEAST}
% [PL: How this all integrates in a flexible framework and publicly available software, one that adequately accounts for phylogenetic uncertainty by averaging over all plausible evolutionary histories for your data
% If there is too little here to build a subsection, you can attach it to the previous subsection and call that 'Bayesian genealogical inference using BEAST'
% ]

Bayesian evolutionary analysis by sampling trees (BEAST) software \citep{Drummond2012}, which is co-developed in our research group is build around the advances described in Subsections~\ref{sub:epidemiology}-\ref{sub:molecular_clock}.

% By assuming either strict or relaxed clock BEAST allows for sampling rooted and time-measured phylogenies,  
%Adopting Bayesian statistical approach, implemented in a flexible framework of molecular evolution \citep{Drummond2012} helps circumvent some of these hindrances by focusing sampling efforts on the topologies 
%% parts of the tree space
%that best fit the data.

>>>>

\subsection{Bayesian evolutionary inference in BEAST}
% [PL: How this all integrates in a flexible framework and publicly available software, one that adequately accounts for phylogenetic uncertainty by averaging over all plausible evolutionary histories for your data
% ]

Bayesian evolutionary analysis by sampling trees (BEAST) software \citep{Drummond2012}, co-developed in our research group, is build around the advances described in Subsections~\ref{sub:epidemiology}-\ref{sub:molecular_clock}.
BEAST is set in a Bayesian statistical framework, meaning that it combines the data likelihood (see Subsection~\ref{sub:likelihood}) with prior knowledge (see Subsection~\ref{sec:bayesian_inference}), and generates posterior samples using MCMC (see Subsection~\ref{sub:MCMC}).

% other packages
BEAST is not the only software package for Bayesian phylogenetic analysis and can be compared to a number of other packages.
MrBayes \citep{Huelsenbeck2001} is a command-line driven software focused on phylogenetic inference.
Migrate-n \citep{migrate} and LAMARC \citep{Kuhner2006} focus on coalescent-driven population genetics.

Like these programs, at the core of the BEAST lies a Metropolis-Hasting algorithm (\cite{Metropolis1953, Hasting1970}, see also Subsection~\ref{sub:mcmc}) for constructing Markov chains that have desired posterior distributions as their equilibrium distribution.
It is precisely why BEAST is not meant for building a single tree or a single estimate of a parameter of interest, and rather generates whole samples that can be used to summarize distributions of mutation rates, divergence times, demographic parameters, trait evolution or others.

%beast is cool
When compared to other similar programs, BEAST definitely stands out in the number of models availiable for performing inference.
Perhaps the most characteristic point is that BEAST is solely focused on sampling rooted, time measured phylogenies.
This allows for a very natural interpretation of reconstructed trees, and is achieved thanks to the adoption of the molecular clock models, from simple strickt clock models, with a uniform rate across the whole time-span of the evolution, to relaxed molecular clock models that model a rate variation \citep{Thorne1998,Yoder2000,Drummond2006,Drummond2010}.
BEAST sports a large number of complementary evolutionary models, which can be combined in a flexible way.
Various substitution models for performing analysis on nucleotide, codon or amino-acid level are complemented with demographic tree priors such as the coalescent, skyline \citep{Drummond2005} or skyride \citep{Minin2008b} models, relaxed clock models and others.

% using beast
The program takes as input structured XML files which describe the data to be analyzed, the models to be used, the number o fposterior samples to collect and the output options.
The output consists of a simple tab-separated files, with each row corresponding to one value sampled from the posterior distribution, for both the parameter values and actual trees.

%gpus
Growing availiability of large complex data-sets has led to the term `Big Data' being coined.
Large sample sizes, such as full genome sequences or the ones generated with th enext-gen sequencing techniques are bound to pose a problem for phylogenetic inference as well.
Fortunately there exist solutions that aim at reducing the time that it takes to fit complex phylogenetic models to the vast amounts of data.
These solutions are mainly focused on harnessing the powers of modern graphics processing units (GPUs), such as the NVIDIAs Tesla line of GPUs that target the high performance computing market.
BEAGLE (Broad-platform evolutionary analysis generalized likelihood evaluator, \cite{Ayres2012}) is a library and an API (Application Programming Interface) that massively parallel graphics processors and other multicore devices to substantially decrease the runtime of phylogenetic analysis \citep{Suchard2009}.
BEAST readily integrates with BEAGLE and provides a GUI interface to configure the library.

% external software
BEAST comes with a number of additional external software packages that can either assist in generating the input for BEAST, analyzing the results or for some other more specific tasks.
\textbf{BEAUti} (Bayesian Evolutionary Analysis Utility) is a  GUI (graphical user-interface) application for generating XML input files for BEAST. 
% It alleviates some of the most cumbersome tasks that would otherwise have to be done manually, thus saving time and simplifying 
It allows for manipulation of models, priors, operators, and MCMC chain settings, without resorting to a manual editing of the XML syntax.
\textbf{LogCombiner} combines log files and tree log files from multiple runs, that can be than importer into \textbf{Tracer}, a graphical tool for the analysis and visualisation of the output of BEAST as well as other common MCMC packages.
It allows for a quick inspection of mixing, convergence, and distribution of the parameters of interest.
\textbf{TreeAnnotator} summarizes the tree log files as MCC (maximum clade credibility) trees and allows for annotating tree branches with external information, such as posterior probabilities, a posteriori most probable ancestral traits and other.
These information can then be visualised using \textbf{FigTree} or if nodes of the trees are annotated with spatial information they can be displayes in geographical coordinates using \textbf{SPREAD}.
Finally \textbf{PIBUSS} is a software for sequence simulation, capable of generating XML files understood by BEAST for combined simulation-inference analysis as well as sequences in formats taht can be imported into any other phylogenetic software.

BEAST is a flexible package for statistical pylogenetics and hypothesis testing.
In addition to a large array of availiable models, it is actively developed with new models and enhancements included with every new stable release.
